---
title: "Creating ADOE"
output:
  rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Creating ADOE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(admiraldev)
```

# Introduction

This article describes creating an ADOE ADaM with Ophthalmology Exam Analysis data for ophthalmology endpoints. It is to be used in conjunction with the article on [creating a BDS dataset from SDTM](https://pharmaverse.github.io/admiral/articles/bds_finding.html). As such, derivations and processes that are not specific to ADOE are absent, and the user is invited to consult the aforementioned article for guidance.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless
otherwise specified.*

## Dataset Contents

`{admiralophtha}` suggests to populate ADOE with general miscellaneous ophthalmology parameters. Any efficacy endpoint-related parameters (eg. BCVA tests) should be placed in separate datasets (eg. ADBCVA).

## Required Packages

The examples of this vignette require the following packages.

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(admiral)
library(pharmaversesdtm)
library(admiraldev)
library(admiralophtha)
library(stringr)
```

# Programming Workflow

* [Initial Set Up of ADOE ](#setup)
* [Assigning `PARAM`/`PARAMCD` and `AVISIT/AVISITN`](#param_avisit)
* [Creating IOP pre to post-dose difference derived parameter](#parameters)
* [Further Derivations of Standard BDS Variables](#further)
* [Example Script](#example)

## Initial set up of ADOE {#setup}

As with all BDS ADaM datasets, one should start from the OE SDTM, where only the general ophthalmology records are of interest. For the purposes of the next two sections, we shall be using the `{admiral}` OE and ADSL test data. We will also require a lookup table for the mapping of parameter codes.

**Note**: to simulate an ophthalmology study, we add a randomly generated `STUDYEYE` variable to ADSL, but in practice `STUDYEYE` will already have been derived using `derive_var_studyeye()`.

```{r}
data("oe_ophtha")
data("admiral_adsl")

# Add STUDYEYE to ADSL to simulate an ophtha dataset
adsl <- admiral_adsl %>%
  as.data.frame() %>%
  mutate(STUDYEYE = sample(c("LEFT", "RIGHT"), n(), replace = TRUE)) %>%
  convert_blanks_to_na()

oe <- convert_blanks_to_na(oe_ophtha)

# Lookup table

# nolint start
param_lookup <- tibble::tribble(
  ~OETESTCD, ~OECAT, ~OESCAT, ~AFEYE, ~PARAMCD, ~PARAM, ~PARAMN,
  "CSUBTH", "OPHTHALMIC ASSESSMENTS", "SD-OCT CST SINGLE FORM", "Study Eye", "SCSUBTH", "Study Eye Center Subfield Thickness (um)", 1,
  "CSUBTH", "OPHTHALMIC ASSESSMENTS", "SD-OCT CST SINGLE FORM", "Fellow Eye", "FCSUBTH", "Fellow Eye Center Subfield Thickness (um)", 2,
  "DRSSR", "OPHTHALMIC ASSESSMENTS", "SD-OCT CST SINGLE FORM", "Study Eye", "SDRSSR", "Study Eye Diabetic Retinopathy Severity", 3,
  "DRSSR", "OPHTHALMIC ASSESSMENTS", "SD-OCT CST SINGLE FORM", "Fellow Eye", "FDRSSR", "Fellow Eye Diabetic Retinopathy Severity", 4,
  "IOP", "INTRAOCULAR PRESSURE", NA_character_, "Study Eye", "SIOP", "Study Eye IOP (mmHg)", 5,
  "IOP", "INTRAOCULAR PRESSURE", NA_character_, "Fellow Eye", "FIOP", "Fellow Eye IOP (mmHg)", 6
)
# nolint end
```

Following this setup, the programmer can start constructing ADOE. The first step is to subset OE to only general ophthalmology parameters. Then, one can merge the resulting dataset with ADSL. This is required for two reasons: firstly, `STUDYEYE` is crucial in the mapping of `AFEYE` and `PARAMCD`'s. Secondly, the treatment start date (`TRTSDT`) is also a prerequisite for the derivation of variables such as Analysis Day (`ADY`).

```{r}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P, STUDYEYE)

adoe <- oe %>%
  filter(
    OETESTCD %in% c("CSUBTH", "DRSSR", "IOP")
  ) %>%
  derive_vars_merged(
    dataset_add = adsl,
    new_vars = adsl_vars,
    by_vars = get_admiral_option("subject_keys")
  )
```

The next item of business is to derive `AVAL`, `AVALU`, and `DTYPE`. In this example, due to the small number of parameters their derivation is trivial. `AFEYE` is also created in this step using the function `derive_var_afeye()`. To determine the affected eye, this function compares `OELAT` to the `STUDYEYE` variable created from the previous step.

```{r}
adoe <- adoe %>%
  # Calculate AVAL, AVALC, AVALU and DTYPE
  mutate(
    AVAL = OESTRESN,
    AVALC = OESTRESC,
    AVALU = OESTRESU,
    DTYPE = NA_character_
  ) %>%
  # Derive AFEYE needed for PARAMCD derivation
  derive_var_afeye(loc_var = OELOC, lat_var = OELAT, loc_vals = c("EYE", "RETINA"))
```

## Assigning `PARAM`/`PARAMCD` and `AVISIT/AVISITN` {#param_avisit}

Moving forwards, `PARAM` and `PARAMCD` can be assigned using `derive_vars_merged()` from `{admiral}` and the lookup table `param_lookup` generated above. `AVISIT`, `AVISITN` and related timepoint variables can also be derived soon after, though their derivation is generally study-specific. A simple option is included below; please consult the `{admiral}` [BDS findings vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html#timing) for a more detailed discussion.

```{r}
adoe <- adoe %>%
  # Add PARAM, PARAMCD from lookup table
  derive_vars_merged(
    dataset_add = param_lookup,
    new_vars = exprs(PARAM, PARAMCD),
    by_vars = exprs(OETESTCD, AFEYE)
  ) %>%
  # Derive visit, baseline flag info and BASETYPE
  mutate(
    ATPTN = OETPTNUM,
    ATPT = OETPT,
    AVISIT = case_when(
      str_detect(VISIT, "SCREEN") ~ "Screening",
      !is.na(VISIT) ~ str_to_title(VISIT),
      TRUE ~ NA_character_
    ),
    AVISITN = round(VISITNUM, 0),
    ABLFL = if_else(AVISIT == "Baseline", "Y", NA_character_)
    # In actual studies, ABLFL derivation will likely be more nuanced
    # and leverage derive_var_extreme_flag()
  )
```

## Creating IOP pre to post-dose difference derived parameter {#parameters}

Two derived parameters of interest are the difference between pre and post-dose IOP in each eye at each visit. These records can be added with two calls to `derive_param_computed()`. Since the calls are very similar, they can be executed in one code block using `call_derivation()` - please see the [Higher Order Functions vignette](https://pharmaverse.github.io/admiral/articles/higher_order.html) for more details. 

```{r}
adoe <- adoe %>%
  # Add derived parameter for difference between pre and post dose IOP
  call_derivation(
    derivation = derive_param_computed,
    by_vars = c(get_admiral_option("subject_keys"), !!adsl_vars, exprs(AVISIT, AVISITN)),
    variable_params = list(
      # Study eye
      params(
        parameters = exprs(
          SIOPPRE = PARAMCD == "SIOP" & ATPT == "PRE-DOSE",
          SIOPPOST = PARAMCD == "SIOP" & ATPT == "POST-DOSE"
        ),
        set_values_to = exprs(
          PARAMCD = "SIOPCHG",
          PARAM = "Study Eye IOP Pre to Post Dose Diff (mmHg)",
          PARAMN = 9,
          AVAL = AVAL.SIOPPOST - AVAL.SIOPPRE,
          AVALC = as.character(AVAL)
        )
      ),
      # Fellow eye
      params(
        parameters = exprs(
          FIOPPRE = PARAMCD == "FIOP" & ATPT == "PRE-DOSE",
          FIOPPOST = PARAMCD == "FIOP" & ATPT == "POST-DOSE"
        ),
        set_values_to = exprs(
          PARAMCD = "FIOPCHG",
          PARAM = "Fellow Eye IOP Pre to Post Dose Diff (mmHg)",
          PARAMN = 10,
          AVAL = AVAL.FIOPPOST - AVAL.FIOPPRE,
          AVALC = as.character(AVAL)
        )
      )
    )
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adoe %>% arrange(USUBJID, AVISIT) %>% select(USUBJID, AVISIT, PARAMCD, AVAL),
  display_vars = exprs(USUBJID, PARAMCD, AVISIT, AVAL),
  filter = str_detect(PARAMCD, "IOP") & USUBJID == "01-701-1028" &
    AVISIT %in% c("Baseline", "Week 4")
)
```

Note that within the call to `derive_param_computed()`, the `parameters` argument has been used to pass an expression that uniquely identifies which records are the pre-dose IOP and which are the post-dose IOP using the timepoint variable `OETPT`, because all IOP records are mapped to `PARAMCD = "SIOP"` or `PARAMCD = "FIOP"`. Users may need to update this expression if their study-specific collection or mapping differs from this standard. 

Additionally, it should be noted that for the `SIOPCHG` and `FIOPCHG` derived parameters, it is generally recommended not to populate `BASE`, `CHG` and `PCHG` as they are difficult/confusing to interpret. This can be simply achieved in one step, as the derivation of `derive_var_base()` can be placed inside of `restrict_derivation()` with a filter added to exclude these parameters. Then, `BASE` will be set to `NA` for `SIOPCHG` and `FIOPCHG`, so later calls to `derive_var_chg()` and `derive_var_pchg()` do not need any changes.

```{r}
adoe <- adoe %>%
  # Calculate BASE (do not derive for IOP change params)
  restrict_derivation(
    derivation = derive_var_base,
    args = params(
      by_vars = c(get_admiral_option("subject_keys"), exprs(PARAMCD, ATPT)),
      source_var = AVAL,
      new_var = BASE
    ),
    filter = !PARAMCD %in% c("SIOPCHG", "FIOPCHG")
  )
```

## Further Derivations of Standard BDS Variables {#further}

The user is invited to consult the article on [creating a BDS dataset from SDTM](https://pharmaverse.github.io/admiral/articles/bds_finding.html) to learn how to add standard BDS variables to ADOE. 

## Example Script {#example}

ADaM | Sample Code 
---- | -------------- 
ADOE | [ad_adoe.R](https://github.com/pharmaverse/admiralophtha/blob/main/inst/templates/ad_adoe.R)

