---
title: "Creating ADVFQ"
output:
  rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Creating ADVFQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
link <- function(text, url) {
  return(
    paste0(
      "[", text, "]",
      "(", url, ")"
    )
  )
}
dyn_link <- function(text,
                     base_url,
                     relative_url = "",
                     # Change to TRUE when admiral adopts multiversion docs
                     is_multiversion = FALSE,
                     multiversion_default_ref = "main") {
  url <- paste(base_url, relative_url, sep = "/")
  if (is_multiversion) {
    url <- paste(
      base_url,
      Sys.getenv("BRANCH_NAME", multiversion_default_ref),
      relative_url,
      sep = "/"
    )
  }
  return(link(text, url))
}
# Other variables
admiral_homepage <- "https://pharmaverse.github.io/admiral"
library(admiraldev)
```

# Introduction

This article describes creating an `ADVFQ` ADaM with Visual Functioning Questionnaire
data for ophthalmology endpoints.

 **Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless
 otherwise specified.* 

## Required Packages

The examples of this vignette require the following packages.

```{r, warning=FALSE, message=FALSE}
library(admiral)
library(admiral.test)
```

# Programming Workflow

* [Read in Data](#readdata) 
* [Derive Dates (`ADT` `ADY`)](#dates)
* [Define Parameters in lookup table](#parameters1)
* [Derive Analysis Value for existing questions (`AVAL`)](#aval)
* [Derive Parameters for recoded items and summary scores](#parameters) 
* [Derive Analysis Visit (`AVISIT`)](#avisit)
* [Derive Baseline variables (`ABLFL` `BASE` `CHG` `PCHG`)](#baseline)
* [Derive Analysis Variables (`ANL01FL` `ASEQ`)](#analysis)
* [Add ADSL Variables](#adslvars) 
* [Example Script](#example) 

## Read in Data {#readdata}

To start, all datasets needed for the creation of the questionnaire dataset
should be read into the environment. This will be a company specific process.

For example purpose, the ADaM datasets---which are included
in `{admiral}` and `{admiralophtha}`---are used.

```{r} 
data("admiral_adsl")
data("admiral_qs")
adsl <- admiral_adsl
qs <- admiral_qs
```

## Derive Dates {#dates}

To derive the date variables `ADT`, `ADY`,
... the `admiral::derive_vars_dt()` function and the `admiral::derive_vars_dy()` function can be
used. This function allows the user to impute the date as well.

```{r, eval=FALSE} 
qs <- qs %>% filter(QSTESTCD %in% c("VFQ1", "VFQ2", "VFQ3", "VFQ4"))

adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)

advfq <- derive_vars_merged(
  qs,
  dataset_add = adsl,
  new_vars = adsl_vars,
  by_vars = exprs(STUDYID, USUBJID)
) %>%
  ## Calculate ADT, ADY ----
  derive_vars_dt(
    new_vars_prefix = "A",
    dtc = QSDTC
  ) %>%
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT))
```

## Derive Parameters in lookup table {#parameters1}
 
Create a parameter lookup table which includes `QSTESTCD`, `PARAMCD`, `PARAM`, `PARCAT1` and `PARCAT2`.
This should include all parameters that will be needed in the final `ADVFQ`. This will be used later to merge Parameter information.

```{r, eval=FALSE} 
param_lookup <- tibble::tribble(
  ~QSTESTCD, ~PARAMCD, ~PARAM, ~PARCAT1, ~PARCAT2,
  "VFQ1", "VFQ1", "Overall Health", "NEI VFQ-25", "Original Response",
  "VFQ2", "VFQ2", "Eyesight in Both Eyes", "NEI VFQ-25", "Original Response",
  "VFQ3", "VFQ3", "Worry About Eyesight", "NEI VFQ-25", "Original Response",
  "VFQ4", "VFQ4", "Pain in and Around Eyes", "NEI VFQ-25", "Original Response",
  "QR01", "QR01", "Recoded Item - 01", "NEI VFQ-25", "General 01",
  "QR02", "QR02", "Recoded Item - 02", "NEI VFQ-25", "General 01",
  "QR03", "QR03", "Recoded Item - 03", "NEI VFQ-25", "General 02",
  "QR04", "QR04", "Recoded Item - 04", "NEI VFQ-25", "General 02",
  "QSG01", "QSG01", "General Score 01", "NEI VFQ-25", "Averaged Result",
  "QSG02", "QSG02", "General Score 02", "NEI VFQ-25", "Averaged Result",
  "QBCSCORE", "QBCSCORE", "Composite Score", "NEI VFQ-25", "Averaged Result"
)
```
## Derive Analysis Value for existing questions {#aval}

To derive the analysis values we use the function `admiral::derive_vars_merged_lookup()` which merges on PARAMCD from the Parameter lookup table.

This merges on the parameter by QSTESCD and assigns AVAL and AVALC.

```{r, eval=FALSE}
advfq <- advfq %>%
  ## Add PARAMCD only - add PARAM etc later ----
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    new_vars = exprs(PARAMCD),
    by_vars = exprs(QSTESTCD)
  ) %>%
  ## Calculate AVAL and AVALC ----
  mutate(
    AVAL = QSSTRESN,
    AVALC = QSORRES
  )
```

## Derive Parameters for recoded items and summary scores {#parameters}

Creates new parameters as new records for recoded existing parameters. 
Run this section of code for every question that you need recoding. This gives an example of recoding one question.

```{r, eval=FALSE}
## QR01 Recoded Item 01
# set to 100 if [advfq.AVAL] = 1
# else set to 75 if [advfq.AVAL] = 2
# else set to 50 if [advfq.AVAL] = 3
# else set to 25 if [advfq.AVAL] = 4
# else set to 0 if [advfq.AVAL] = 5
advfq <- advfq %>%
  derive_summary_records(
    by_vars = exprs(STUDYID, USUBJID, !!!adsl_vars, PARAMCD, VISITNUM, VISIT, ADT, ADY),
    filter = QSTESTCD == "VFQ1" & !is.na(AVAL),
    analysis_var = AVAL,
    summary_fun = identity,
    set_values_to = exprs(PARAMCD = "QR01")
  ) %>%
  mutate(AVAL = ifelse(PARAMCD == "QR01",
    case_when(
      AVAL == 1 ~ 100,
      AVAL == 2 ~ 75,
      AVAL == 3 ~ 50,
      AVAL == 4 ~ 25,
      AVAL >= 5 ~ 0
    ),
    AVAL
  ))
```

To create summary records as average of recoded questions. This example uses two of the recoded questions.

```{r, eval=FALSE}
## Derive a new record as a summary record  ----
## QSG01 General Score 01
# Average of QR01 and QR02 records
advfq <- advfq %>%
  derive_summary_records(
    by_vars = exprs(STUDYID, USUBJID, !!!adsl_vars, VISITNUM, VISIT, ADT, ADY),
    filter = PARAMCD %in% c("QR01", "QR02") & !is.na(AVAL),
    analysis_var = AVAL,
    summary_fun = mean,
    set_values_to = exprs(PARAMCD = "QSG01")
  )
```

## Derive Analysis Visit {#avisit}

Derives AVISIT and AVISITN based on VISIT. See also the [`Visit and Period Variables`](https://pharmaverse.github.io/admiral/articles/visits_periods.html#visits) vignette. 


```{r, eval=FALSE}
advfq <- advfq %>%
  # Derive Timing
  mutate(
    AVISIT = case_when(
      str_detect(VISIT, "SCREEN|UNSCHED|RETRIEVAL|AMBUL") ~ NA_character_,
      # If VISIT=DAY 1 then set to Baseline, study specific
      str_detect(VISIT, "DAY 1") ~ "Baseline",
      !is.na(VISIT) ~ str_to_title(VISIT),
      TRUE ~ NA_character_
    ),
    AVISITN = as.numeric(case_when(
      VISIT == "BASELINE" ~ "0",
      str_detect(VISIT, "WEEK") ~ str_trim(str_replace(VISIT, "WEEK", "")),
      TRUE ~ NA_character_
    ))
  )
```

## Derive Baseline variables {#baseline}

Here we derive ABLFL, BASE, CHG and PCHG variables.

```{r, eval=FALSE}
## Derive baseline flags ----
advfq <- advfq %>%
  # Calculate ABLFL
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order = exprs(ADT, VISITNUM, QSSEQ),
      new_var = ABLFL,
      mode = "last"
    ),
    filter = (!is.na(AVAL) &
      ADT <= TRTSDT)
  )

## Derive baseline information ----
advfq <- advfq %>%
  # Calculate BASE
  derive_var_base(
    by_vars = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = AVAL,
    new_var = BASE
  ) %>%
  # Calculate CHG
  derive_var_chg() %>%
  # Calculate PCHG
  derive_var_pchg()
```
## Derive Analysis Variables {#analysis}

In most finding ADaMs, an analysis flag is derived to identify the appropriate observation(s) to use for a particular analysis when a subject has multiple observations within a particular timing period.

In this situation, an analysis flag (e.g. `ANLxxFL`) may be used to choose the appropriate record for analysis.

This flag may be derived using the [`admiral`](https://pharmaverse.github.io/admiral/) function `admiral::derive_var_extreme_flag()`. For this example, we will assume we would like to choose the latest value by USUBJID, PARAMCD and AVISIT.

```{r, eval=FALSE}
## ANL01FL: Flag last result within an AVISIT for post-baseline records ----
advfq <- advfq %>%
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      new_var = ANL01FL,
      by_vars = exprs(USUBJID, PARAMCD, AVISIT),
      order = exprs(ADT, AVAL),
      mode = "last"
    ),
    filter = !is.na(AVISITN) & ONTRTFL == "Y"
  )
```

We then derive ASEQ using `admiral::derive_var_obs_number()` based on the observation number within the dataset, additionally merge on `PARAM`, `PARCAT1` and `PARCAT2` using the earlier lookup table.

```{r, eval=FALSE}
## Get ASEQ and PARAM  ----
advfq <- advfq %>%
  # Calculate ASEQ
  derive_var_obs_number(
    new_var = ASEQ,
    by_vars = exprs(STUDYID, USUBJID),
    order = exprs(PARAMCD, ADT, AVISITN, VISITNUM),
    check_type = "error"
  ) %>%
  # Derive PARAM
  derive_vars_merged(dataset_add = select(param_lookup, -QSTESTCD), by_vars = exprs(PARAMCD))
```
## Add ADSL Variables {#adslvars} 

Variables from ADSL which are required.

```{r, eval=FALSE}
# Add all ADSL variables
advfq <- advfq %>%
  derive_vars_merged(
    dataset_add = select(adsl, !!!negate_vars(adsl_vars)),
    by_vars = exprs(STUDYID, USUBJID)
  )
```
## Example Script {#example}

ADaM | Sample Code 
 ---- | -------------- 
ADVFQ | `r dyn_link("ad_advfq.R", "https://github.com/pharmaverse/admiralophtha/blob", "inst/templates/ad_advfq.R", is_multiversion = TRUE)`
