---
title: "Creating ADVFQ"
output:
  rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Creating ADVFQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
link <- function(text, url) {
  return(
    paste0(
      "[", text, "]",
      "(", url, ")"
    )
  )
}
dyn_link <- function(text,
                     base_url,
                     relative_url = "",
                     # Change to TRUE when admiral adopts multiversion docs
                     is_multiversion = FALSE,
                     multiversion_default_ref = "main") {
  url <- paste(base_url, relative_url, sep = "/")
  if (is_multiversion) {
    url <- paste(
      base_url,
      Sys.getenv("BRANCH_NAME", multiversion_default_ref),
      relative_url,
      sep = "/"
    )
  }
  return(link(text, url))
}
# Other variables
admiral_homepage <- "https://pharmaverse.github.io/admiral"
library(admiraldev)
```

# Introduction

This article describes creating an `ADVFQ` ADaM with Visual Functioning Questionnaire
data for ophthalmology endpoints.

 **Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless
 otherwise specified.* 

## Required Packages

The examples of this vignette require the following packages.

```{r, warning=FALSE, message=FALSE}
library(admiral)
library(admiral.test)
```

# Programming Workflow

* [Read in Data](#readdata) 
* [Derive Dates (`ADT` `ADY`)](#dates)
* [Define Parameters in lookup table](#parameters1)
* [Derive Analysis Value for existing questions (`AVAL`)](#aval)
* [Derive Parameters for recoded items and summary scores](#parameters) 
* [Derive Analysis Visit (`AVISIT`)](#avisit)
* [Derive Baseline variables (`ABLFL` `BASE` `CHG` `PCHG`)](#baseline)
* [Derive Analysis Variables (`ANL01FL` `ASEQ`)](#analysis)
* [Add ADSL Variables](#adslvars) 
* [Example Script](#example) 

## Read in Data {#readdata}

To start, all datasets needed for the creation of the questionnaire dataset
should be read into the environment. This will be a company specific process.

For example purpose, the ADaM datasets---which are included
in `{admiral}` and `{admiralophtha}`---are used.

```{r} 
data("admiral_adsl")
data("admiral_qs")
adsl <- admiral_adsl
qs <- admiral_qs
```

## Derive Dates {#dates}

To derive the date variables `ADT`, `ADY`,
... the `admiral::derive_vars_dt()` function and the `admiral::derive_vars_dy()` function can be
used. This function allows the user to impute the date as well.

```{r, eval=FALSE} 
adsl_vars <- vars(TRTSDT, TRTEDT, TRT01A, TRT01P)

adqs <- derive_vars_merged(
  qs,
  dataset_add = adsl,
  new_vars = adsl_vars,
  by_vars = vars(STUDYID, USUBJID)
) %>%
  ## Calculate ADT, ADY ----
derive_vars_dt(
  new_vars_prefix = "A",
  dtc = QSDTC
) %>%
derive_vars_dy(reference_date = TRTSDT, source_vars = vars(ADT))
```

## Derive Parameters in lookup table {#parameters1}
 
Create a parameter lookup table which includes `QSTESTCD`, `PARAMCD`, `PARAM`, `PARCAT1` and `PARCAT2`.
This should include all parameters that will be needed in the final `ADVFQ`. This will be used later to merge Parameter information.

```{r, eval=FALSE} 
param_lookup <- tibble::tribble(
  ~QSTESTCD, ~PARAMCD, ~PARAM,~PARCAT1, ~PARCAT2,
  "ACITM01", "ACITM01", "WORD RECALL", "NEI VFQ 25", "Original Response",
  "ACITM02", "ACITM02", "NAMING OBJECTS", "NEI VFQ 25", "Original Response",
  "ACITM03", "ACITM03", "DELAYED WORD RECALL", "NEI VFQ 25", "Original Response",
  "ACITM04", "ACITM04", "COMMANDS", "NEI VFQ 25", "Original Response",
  "QR01", "QR01", "Recoded Item - 01", "NEI VFQ 25", "General 01",
  "QR02", "QR02", "Recoded Item - 02", "NEI VFQ 25", "General 01",
  "QR03", "QR03", "Recoded Item - 03", "NEI VFQ 25", "General 02",
  "QR04", "QR04", "Recoded Item - 04", "NEI VFQ 25", "General 02",
  "QSG01", "QSG01", "General Score 01", "NEI VFQ 25", "Averaged Result",
  "QSG02", "QSG02", "General Score 02", "NEI VFQ 25", "Averaged Result",
  "QBCSCORE", "QBCSCORE", "Composite Score", "NEI VFQ 25", "Averaged Result"
) 
```
## Derive Analysis Value for existing questions {#aval}

To derive the analysis values we use the function `admiral::derive_vars_merged_lookup()` which merges on PARAMCD from the Parameter lookup table.

This merges on the parameter by QSTESCD and assigns AVAL and AVALC.

```{r, eval=FALSE}
adqs <- adqs %>%
  ## Add PARAMCD only - add PARAM etc later ----
derive_vars_merged_lookup(
  dataset_add = param_lookup,
  new_vars = vars(PARAMCD),
  by_vars = vars(QSTESTCD)
) %>%
  ## Calculate AVAL and AVALC ----
mutate(
  AVAL = QSSTRESN,
  AVALC = QSORRES
)
```

## Derive Parameters for recoded items and summary scores {#parameters}

Creates new parameters as new records for recoded existing parameters. 
Run this section of code for every question that you need recoding. This gives an example of recoding one question.

```{r, eval=FALSE}
## QR01 Recoded Item 01
# set to 100 if [ADQS.AVAL] = 1
# else set to 75 if [ADQS.AVAL] = 2
# else set to 50 if [ADQS.AVAL] = 3
# else set to 25 if [ADQS.AVAL] = 4
# else set to 0 if [ADQS.AVAL] = 5
adqs <- adqs %>%
  derive_summary_records(
    by_vars = vars(STUDYID, USUBJID, !!!adsl_vars, PARAMCD, VISITNUM, VISIT, ADT, ADY),
    filter = QSTESTCD=="ACITM01" & !is.na(AVAL),
    analysis_var = AVAL,
    summary_fun = identity,
    set_values_to = vars(PARAMCD='QR01')
  ) %>%
  mutate(AVAL= ifelse(PARAMCD=="QR01",
                        case_when(AVAL==1 ~ 100,
                        AVAL==2 ~ 75,
                        AVAL==3 ~ 50,
                        AVAL==4 ~ 25,
                        AVAL>=5 ~ 0),
                        AVAL))
```

To create summary records as average of recoded questions. This example uses two of the recoded questions.

```{r, eval=FALSE}
## Derive a new record as a summary record  ----
## QSG01 General Score 01
# Average of QR01 and QR02 records
adqs <- adqs %>%
  derive_summary_records(
    by_vars = vars(STUDYID, USUBJID, !!!adsl_vars, VISITNUM, VISIT, ADT, ADY),
    filter = PARAMCD %in% c("QR01", "QR02") & !is.na(AVAL),
    analysis_var = AVAL,
    summary_fun = mean,
    set_values_to = vars(PARAMCD="QSG01")
  )
```

## Derive Analysis Visit {#avisit}

Derives AVISIT and AVISITN based on VISIT. See also the [`Visit and Period Variables`](https://pharmaverse.github.io/admiral/articles/visits_periods.html#visits) vignette. 


```{r, eval=FALSE}
adqs <- adqs %>%
  # Derive Timing
  mutate(
    AVISIT = case_when(
      str_detect(VISIT, "SCREEN|UNSCHED|RETRIEVAL|AMBUL") ~ NA_character_,
      #If VISIT=DAY 1 then set to Baseline, study specific
      str_detect(VISIT, "DAY 1") ~ "Baseline",
      !is.na(VISIT) ~ str_to_title(VISIT),
      TRUE ~ NA_character_
    ),
    AVISITN = as.numeric(case_when(
      VISIT == "BASELINE" ~ "0",
      str_detect(VISIT, "WEEK") ~ str_trim(str_replace(VISIT, "WEEK", "")),
      TRUE ~ NA_character_
    ))
  )
```

## Derive Baseline variables {#baseline}

Here we derive ABLFL, BASE, CHG and PCHG variables.

```{r, eval=FALSE}
## Derive baseline flags ----
adqs <- adqs %>%
 # Calculate ABLFL
 restrict_derivation(
 derivation = derive_var_extreme_flag,
 args = params(
 by_vars = vars(STUDYID, USUBJID, PARAMCD),
 order = vars(ADT, VISITNUM, QSSEQ),
 new_var = ABLFL,
 mode = "last"
 ),
 filter = (!is.na(AVAL) &
 ADT <= TRTSDT)
)

## Derive baseline information ----
adqs <- adqs %>%
 # Calculate BASE
 derive_var_base(
 by_vars = vars(STUDYID, USUBJID, PARAMCD),
 source_var = AVAL,
 new_var = BASE
 ) %>%
 # Calculate CHG
 derive_var_chg() %>%
 # Calculate PCHG
 derive_var_pchg()
```
## Derive Analysis Variables {#analysis}

In most finding ADaMs, an analysis flag is derived to identify the appropriate observation(s) to use for a particular analysis when a subject has multiple observations within a particular timing period.

In this situation, an analysis flag (e.g. `ANLxxFL`) may be used to choose the appropriate record for analysis.

This flag may be derived using the [`admiral`](https://pharmaverse.github.io/admiral/) function `admiral::derive_var_extreme_flag()`. For this example, we will assume we would like to choose the latest value by USUBJID, PARAMCD and AVISIT.

```{r, eval=FALSE}
## ANL01FL: Flag last result within an AVISIT for post-baseline records ----
adqs <- adqs %>%
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      new_var = ANL01FL,
      by_vars = vars(USUBJID, PARAMCD, AVISIT),
      order = vars(ADT, AVAL),
      mode = "last"
    ),
    filter = !is.na(AVISITN) & ONTRTFL == "Y"
  )
```

We then derive ASEQ using `admiral::derive_var_obs_number()` based on the observation number within the dataset, additionally merge on `PARAM`, `PARCAT1` and `PARCAT2` using the earlier lookup table.

```{r, eval=FALSE}
## Get ASEQ and PARAM  ----
adqs <- adqs %>%
  # Calculate ASEQ
  derive_var_obs_number(
    new_var = ASEQ,
    by_vars = vars(STUDYID, USUBJID),
    order = vars(PARAMCD, ADT, AVISITN, VISITNUM),
    check_type = "error"
  ) %>%
  # Derive PARAM
  derive_vars_merged(dataset_add = select(param_lookup, -QSTESTCD), by_vars = vars(PARAMCD))
```
## Add ADSL Variables {#adslvars} 

Variables from ADSL which are required.

```{r, eval=FALSE}
# Add all ADSL variables
adqs <- adqs %>%
  derive_vars_merged(
    dataset_add = select(adsl, !!!negate_vars(adsl_vars)),
    by_vars = vars(STUDYID, USUBJID)
  )
```
## Example Script {#example}

ADaM | Sample Code 
 ---- | -------------- 
ADVFQ | `r dyn_link("ad_advfq.R", "https://github.com/pharmaverse/admiralophtha/blob", "inst/templates/ad_advfq.R", is_multiversion = TRUE)`
